---
title: "MISY262 presentation by group 2"
output:
  html_document:
    df_print: paged
---

---

### Credit Card Default Project - Full EDA + Multiple Models

This script:
- Loads the UCI credit card default dataset
- Performs basic EDA and descriptive statistics
- Explores and (optionally) imputes missing data
- Preprocesses variables (factors, target, etc.)
- Trains multiple ML models for default prediction: Logistic Regression, Decision Tree, Random Forest, XGBoost, Cost-sensitive ensemble (custom "new" method)
- Compares their performance

---

# 0. Setup environment #

```{r}
rm(list = ls())  # clear workspace

# Optional: set working directory where UCI_Credit_Card.csv is stored
# setwd("YOUR/PATH/HERE")

# Load required packages
# If not installed, run install.packages("package_name")
library(dplyr)        # data manipulation
library(ggplot2)      # basic plots
library(caret)        # train/test split and evaluation
library(pROC)         # ROC and AUC
library(rpart)        # decision tree
library(randomForest) # random forest
library(xgboost)      # gradient boosting
library(naniar)       # missing-data visualization
library(mice)      # multiple imputation by chained equations
```

# 1. Load the dataset  #

```{r}
# Read the UCI credit card default dataset
credit <- read.csv("UCI_Credit_Card.csv")
```

# 2. Basic EDA: structure and descriptive stats #

```{r}
# Basic information about the dataset
n_rows <- nrow(credit)
n_cols <- ncol(credit)

cat("Number of rows:", n_rows, "\n")
cat("Number of columns:", n_cols, "\n")

# Display variable names
cat("Variable names:\n")
print(names(credit))

# Peek at the first few rows
head(credit)

# Structure of the dataset: variable types and examples
str(credit)

# Basic summary statistics (for all variables)
summary(credit)

# Example: descriptive statistics for a key financial variable
# LIMIT_BAL is the amount of the given credit (NT dollar)
summary(credit$LIMIT_BAL)

# Distribution of the target variable (0/1 default flag)
# Original target name in this dataset is "default.payment.next.month"
table(credit$default.payment.next.month)
prop.table(table(credit$default.payment.next.month))

```

# 3. Missing data exploration (based on course EDA slides) #

```{r}
# Overall number of missing values in the dataset
total_na <- sum(is.na(credit))
cat("Total number of missing values in the dataset:", total_na, "\n")

# Number of missing values per variable
na_per_col <- colSums(is.na(credit))
cat("Number of missing values per column:\n")
print(na_per_col)

# Visualization of missing data patterns using naniar
# This will create a heatmap of missingness
vis_miss(credit)

# This will show percentage of missing per variable
gg_miss_var(credit, show_pct = TRUE)

# Note:
# In the original UCI credit card dataset, there are typically no NA values.
# If your version of the dataset has no missing data, the above outputs
# will show zero NAs. In that case, data imputation is not required and
# you can explicitly state this in your report.
```

# 4. Missing-data imputation framework (mice) #

```{r}
# We use the result from Task 03:

# total_na = total number of missing values in "credit".

# If there are no missing values, we skip the imputation step.

if (total_na == 0) {
cat("There are no missing values in this dataset, so no imputation is performed.\n")

# For consistency in later code, we simply set:

credit_imputed <- credit

} else {
cat("Some variables contain missing values. We illustrate multiple imputation using the mice package.\n")

# Visualize the missing-data pattern

md.pattern(credit)   # table + pattern plot

# Run multiple imputation by chained equations

imputed <- mice(
credit,
m      = 3,        # number of imputed datasets
method = "pmm",    # predictive mean matching for numeric variables
maxit  = 5         # number of iterations
)

# Choose one completed dataset (for example, dataset 2)

credit_imputed <- complete(imputed, 2)

# From this point on, you could continue the analysis using

# 'credit_imputed' instead of 'credit'.

}

# In the UCI_Credit_Card dataset used in this project, there are typically

# no missing values, so credit_imputed == credit in practice.

```

# 5. Variable type preprocessing #

```{r}
# Drop ID (identifier, not useful for prediction)
credit <- credit %>% dplyr::select(-ID)

# Rename the target variable for convenience
colnames(credit)[colnames(credit) == "default.payment.next.month"] <- "default_next_month"

# Convert target to factor: 0 -> no_default, 1 -> default
credit$default_next_month <- factor(
  credit$default_next_month,
  levels = c(0, 1),
  labels = c("no_default", "default")
)

# Define variables that should be treated as categorical
categorical_vars <- c(
  "SEX",
  "EDUCATION",
  "MARRIAGE",
  "PAY_0",
  "PAY_2",
  "PAY_3",
  "PAY_4",
  "PAY_5",
  "PAY_6"
)

# Cast selected variables to factor
credit[categorical_vars] <- lapply(credit[categorical_vars], factor)

# Check the updated structure
str(credit)


```

# 6. Train/test split for classification (PD)    #

```{r}
set.seed(123)  # for reproducibility

# Create indices for the training set (70% of the data)
train_index <- createDataPartition(
  credit$default_next_month,
  p   = 0.7,
  list = FALSE
)

# Split the data into training and test sets
train_data <- credit[train_index, ]
test_data  <- credit[-train_index, ]

# Make sure factor levels are consistent between train and test
# This avoids errors such as "new levels" in the test set
# (e.g., PAY_4 having a level in test_data that was not seen in train_data).

# IMPORTANT: this uses the 'categorical_vars' vector defined
# in the variable preprocessing section.
for (v in categorical_vars) {
  # Drop unused levels in training data
  train_data[[v]] <- droplevels(train_data[[v]])
  
  # Force test_data to use exactly the same levels as train_data.
  # Any extra levels in test_data will be converted to NA.
  test_data[[v]] <- factor(test_data[[v]], levels = levels(train_data[[v]]))
}

# Align target factor levels as well (usually already consistent)
train_data$default_next_month <- droplevels(train_data$default_next_month)
test_data$default_next_month  <- factor(
  test_data$default_next_month,
  levels = levels(train_data$default_next_month)
)

# Create numeric labels for models that require 0/1 (e.g., XGBoost)
train_label <- ifelse(train_data$default_next_month == "default", 1, 0)
test_label  <- ifelse(test_data$default_next_month == "default", 1, 0)

# Quick sanity check
table(train_data$default_next_month)
table(test_data$default_next_month)

```

# 7. Model 1: Logistic Regression (baseline PD) #

```{r}
logit_model <- glm(
  default_next_month ~ .,
  data   = train_data,
  family = binomial(link = "logit")
)

summary(logit_model)

# Predicted probabilities on test data
probs_logit <- predict(
  logit_model,
  newdata = test_data,
  type    = "response"
)

# Class predictions with 0.5 cutoff
pred_logit_class <- ifelse(probs_logit >= 0.5, "default", "no_default")
pred_logit_class <- factor(pred_logit_class, levels = c("no_default", "default"))

# Confusion matrix and metrics
cm_logit <- confusionMatrix(
  data      = pred_logit_class,
  reference = test_data$default_next_month,
  positive  = "default"
)
cm_logit

# ROC and AUC for logistic regression
roc_logit <- roc(
  response  = test_data$default_next_month,
  predictor = probs_logit,
  levels    = c("no_default", "default"),
  direction = "<"
)
auc_logit <- auc(roc_logit)
auc_logit

```

# 8. Model 2: Decision Tree (rpart) #

```{r}
tree_model <- rpart(
  default_next_month ~ .,
  data   = train_data,
  method = "class"
)

# Predicted probabilities for "default"
probs_tree <- predict(
  tree_model,
  newdata = test_data,
  type   = "prob"
)[, "default"]

# Class predictions with 0.5 cutoff
pred_tree_class <- ifelse(probs_tree >= 0.5, "default", "no_default")
pred_tree_class <- factor(pred_tree_class, levels = c("no_default", "default"))

cm_tree <- confusionMatrix(
  data      = pred_tree_class,
  reference = test_data$default_next_month,
  positive  = "default"
)
cm_tree

roc_tree <- roc(
  response  = test_data$default_next_month,
  predictor = probs_tree,
  levels    = c("no_default", "default"),
  direction = "<"
)
auc_tree <- auc(roc_tree)
auc_tree
```

# 9. Model 3: Random Forest #

```{r}
set.seed(123)

rf_model <- randomForest(
  default_next_month ~ .,
  data       = train_data,
  ntree      = 300,
  mtry       = 6,
  importance = TRUE
)

# Predicted probabilities for "default"
probs_rf <- predict(
  rf_model,
  newdata = test_data,
  type   = "prob"
)[, "default"]

pred_rf_class <- ifelse(probs_rf >= 0.5, "default", "no_default")
pred_rf_class <- factor(pred_rf_class, levels = c("no_default", "default"))

cm_rf <- confusionMatrix(
  data      = pred_rf_class,
  reference = test_data$default_next_month,
  positive  = "default"
)
cm_rf

roc_rf <- roc(
  response  = test_data$default_next_month,
  predictor = probs_rf,
  levels    = c("no_default", "default"),
  direction = "<"
)
auc_rf <- auc(roc_rf)
auc_rf

```

# 10. Model 4: XGBoost #

```{r}
###########################################
# Task 10: XGBoost (NA-safe implementation)
###########################################

# For XGBoost, we create dedicated copies of the data
# and drop any rows that contain NA values. This ensures
# that the design matrices and label vectors have the
# same number of rows.

# Create NA-free versions for XGBoost
train_xgb <- na.omit(train_data)
test_xgb  <- na.omit(test_data)

cat("Number of rows in train_data (original):", nrow(train_data), "\n")
cat("Number of rows in train_xgb (NA-free):", nrow(train_xgb), "\n")

cat("Number of rows in test_data (original):", nrow(test_data), "\n")
cat("Number of rows in test_xgb (NA-free):", nrow(test_xgb), "\n")

# Build model matrices with one-hot encoding for categorical variables
train_matrix <- model.matrix(
  default_next_month ~ . - 1,
  data = train_xgb
)
test_matrix <- model.matrix(
  default_next_month ~ . - 1,
  data = test_xgb
)

# Create numeric labels (0/1) that match these NA-free data sets
train_label <- ifelse(train_xgb$default_next_month == "default", 1, 0)
test_label  <- ifelse(test_xgb$default_next_month == "default", 1, 0)

# Sanity check: the number of labels must equal the number of rows
stopifnot(length(train_label) == nrow(train_matrix))
stopifnot(length(test_label)  == nrow(test_matrix))

# Build DMatrix objects for XGBoost
dtrain <- xgb.DMatrix(
  data  = train_matrix,
  label = train_label
)
dtest <- xgb.DMatrix(
  data  = test_matrix,
  label = test_label
)

# Define XGBoost parameters for binary classification
params <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  max_depth        = 4,
  eta              = 0.1,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

set.seed(123)
xgb_model <- xgb.train(
  params  = params,
  data    = dtrain,
  nrounds = 100,
  verbose = 0
)

# Predicted probabilities from XGBoost
probs_xgb <- predict(xgb_model, newdata = dtest)

# Class predictions using 0.5 cutoff
pred_xgb_class <- ifelse(probs_xgb >= 0.5, "default", "no_default")
pred_xgb_class <- factor(pred_xgb_class, levels = c("no_default", "default"))

# Confusion matrix
cm_xgb <- confusionMatrix(
  data      = pred_xgb_class,
  reference = test_xgb$default_next_month,  # use the NA-free test labels
  positive  = "default"
)
cm_xgb

# ROC and AUC for XGBoost
roc_xgb <- roc(
  response  = test_xgb$default_next_month,
  predictor = probs_xgb,
  levels    = c("no_default", "default"),
  direction = "<"
)
auc_xgb <- auc(roc_xgb)
auc_xgb

```

# 11. Model 5 (NEW): Cost-sensitive ensemble decision rule #

```{r}
# We build the ensemble on the same test set that is used for XGBoost,
# i.e., the NA-free test_xgb. This guarantees that all probability
# vectors and label vectors have the same length.

# Recompute probabilities from logistic regression and random forest
# on test_xgb, so that they are aligned with probs_xgb.

# Logistic regression probabilities on test_xgb
probs_logit_ens <- predict(
  logit_model,
  newdata = test_xgb,
  type    = "response"
)

# Random forest probabilities on test_xgb
probs_rf_ens <- predict(
  rf_model,
  newdata = test_xgb,
  type    = "prob"
)[, "default"]

# XGBoost probabilities on test_xgb were already computed as probs_xgb
# in the XGBoost task, and they have length nrow(test_xgb).

# Sanity check: all three vectors must have the same length
stopifnot(
  length(probs_logit_ens) == nrow(test_xgb),
  length(probs_rf_ens)    == nrow(test_xgb),
  length(probs_xgb)       == nrow(test_xgb)
)

# Ensemble probability as the simple average of the three models
ensemble_probs <- (probs_logit_ens + probs_rf_ens + probs_xgb) / 3

# Numeric labels (0/1) for cost-based threshold selection,
# aligned with test_xgb
ens_label <- ifelse(test_xgb$default_next_month == "default", 1, 0)

# Cost parameters:
#   cost_fn = cost of a false negative (missed default)
#   cost_fp = cost of a false positive (rejecting a good client)
cost_fn <- 10
cost_fp <- 1

# Cost function for a given threshold
compute_cost <- function(threshold, probs, true_labels, cost_fn, cost_fp) {
  # 1 = default, 0 = no_default
  pred <- ifelse(probs >= threshold, 1, 0)
  true <- true_labels
  
  tp <- sum(pred == 1 & true == 1)
  fp <- sum(pred == 1 & true == 0)
  fn <- sum(pred == 0 & true == 1)
  tn <- sum(pred == 0 & true == 0)
  
  total_cost <- cost_fn * fn + cost_fp * fp
  
  return(list(
    total_cost = total_cost,
    tp = tp, fp = fp, fn = fn, tn = tn
  ))
}

# Grid search for the best threshold on the ensemble probabilities
threshold_grid <- seq(0.1, 0.9, by = 0.01)

cost_results <- data.frame(
  threshold  = threshold_grid,
  total_cost = NA_real_
)

for (i in seq_along(threshold_grid)) {
  th  <- threshold_grid[i]
  res <- compute_cost(
    threshold   = th,
    probs       = ensemble_probs,
    true_labels = ens_label,
    cost_fn     = cost_fn,
    cost_fp     = cost_fp
  )
  cost_results$total_cost[i] <- res$total_cost
}

# Find the threshold that minimizes the total expected cost
best_index     <- which.min(cost_results$total_cost)
best_threshold <- cost_results$threshold[best_index]
best_threshold

# Final class predictions using the optimal threshold
pred_ens_class <- ifelse(ensemble_probs >= best_threshold, "default", "no_default")
pred_ens_class <- factor(pred_ens_class, levels = c("no_default", "default"))

# Confusion matrix for the ensemble method
cm_ens <- confusionMatrix(
  data      = pred_ens_class,
  reference = test_xgb$default_next_month,  # use test_xgb, not test_data
  positive  = "default"
)
cm_ens

# ROC and AUC for the ensemble probabilities
roc_ens <- roc(
  response  = test_xgb$default_next_month,
  predictor = ensemble_probs,
  levels    = c("no_default", "default"),
  direction = "<"
)
auc_ens <- auc(roc_ens)
auc_ens

```
# 12: Save ROC curves as PNG figures #

```{r}
# 1) Logistic regression ROC
png("roc_logit.png", width = 800, height = 600)
plot(roc_logit,
     main = "ROC Curve - Logistic Regression")
dev.off()

# 2) Decision tree ROC
png("roc_tree.png", width = 800, height = 600)
plot(roc_tree,
     main = "ROC Curve - Decision Tree")
dev.off()

# 3) Random forest ROC
png("roc_rf.png", width = 800, height = 600)
plot(roc_rf,
     main = "ROC Curve - Random Forest")
dev.off()

# 4) XGBoost ROC
png("roc_xgb.png", width = 800, height = 600)
plot(roc_xgb,
     main = "ROC Curve - XGBoost")
dev.off()

# 5) Ensemble ROC
png("roc_ens.png", width = 800, height = 600)
plot(roc_ens,
     main = "ROC Curve - Cost-sensitive Ensemble")
dev.off()
```


# Summary (for report writing, not code output):

EDA section:
- Basic structure, summary statistics, and distribution of LIMIT_BAL
- Class balance of default_next_month
- Missing-data analysis (total NAs, per-variable NAs, vis_miss, gg_miss_var)
- Conclusion: UCI_Credit_Card dataset has no (or very few) missing values, so no imputation was necessary (or describe the chosen method).

Modeling section:
- Five methods:(1) Logistic Regression (2) Decision Tree (3) Random Forest(4) XGBoost (5) Cost-sensitive ensemble (proposed method)
- Compare AUC, accuracy, and cost-sensitive performance.
- Interpret important variables from logistic / tree / RF models.
